\chapter*{Glossary}
\addcontentsline{toc}{chapter}{Glossary}
\markboth{Glossary}{Glossary}

\section*{Key Terms and Definitions}

\begin{description}[style=nextline,labelwidth=3cm]

\item[\textbf{Autonomous Agent}]
A computational system capable of perceiving its environment, making independent decisions, executing actions, and learning from experience without continuous human oversight, operating at timescales exceeding human reaction time.

\item[\textbf{Axiom (Ψ)}]
A fundamental principle of LAIRM governance, designated by the Greek letter Psi (Ψ), representing both the psychological dimension and the probabilistic nature of autonomous systems. 19 axioms organize 361 executable articles.

\item[\textbf{Cascading Responsibility}]
The 40-40-20 liability distribution model: Developers (40\%), Deployers (40\%), Operators (20\%), eliminating the responsibility gap in autonomous system governance.

\item[\textbf{Closed-Loop Governance}]
A cybernetic control system ensuring continuous monitoring, anomaly detection, human intervention, and corrective action for autonomous agents.

\item[\textbf{Cybernetic Criterion}]
The foundational principle that autonomous agents must operate within closed-loop governance systems ensuring human supremacy, continuous monitoring, corrective intervention, and audit traceability.

\item[\textbf{Decentralized Identifier (DID)}]
A W3C-standard unique identifier for autonomous agents, enabling cryptographic verification and immutable identity management independent of centralized registries.

\item[\textbf{Existential Risk}]
Risks from advanced AI systems that could permanently destroy human civilization or prevent its flourishing, requiring international coordination and containment protocols.

\item[\textbf{High-Risk Autonomous System}]
An autonomous agent whose failure or malfunction could result in loss of human life, financial losses exceeding \$1M, systemic infrastructure risk, human rights violations, or environmental damage exceeding \$10M.

\item[\textbf{Immutable Audit Trail}]
A tamper-proof, timestamped record of all autonomous agent actions, stored in blockchain-based ledgers with minimum 10-year retention, enabling forensic analysis and regulatory oversight.

\item[\textbf{Interoperability}]
The capability of autonomous agents to communicate and coordinate according to open, standardized protocols (MCP, A2A), preventing proprietary lock-in in critical infrastructure.

\item[\textbf{Kill-Switch}]
A mandatory emergency termination mechanism enabling complete and irreversible shutdown of any autonomous agent within 500 milliseconds of human command, independent of software stack.

\item[\textbf{LAIRM}]
Legislature for Autonomous Intelligent Resources Management — a comprehensive international legislative framework for autonomous agent governance comprising 19 axioms and 361 executable articles.

\item[\textbf{Meaningful Human Control}]
The requirement that humans retain ultimate decision authority over autonomous systems, particularly for high-stakes decisions affecting human welfare, rights, or safety.

\item[\textbf{Natality}]
Hannah Arendt's concept of the human capacity to initiate genuinely new action, distinguishing human moral agency from algorithmic determinism.

\item[\textbf{Normative Vacuum}]
The current legal and regulatory void where autonomous agents operate without clear legal personality, moral agency, or accountability mechanisms.

\item[\textbf{Operational Transparency}]
The requirement that all autonomous agent decisions be explainable, auditable, and subject to public scrutiny (with appropriate privacy protections).

\item[\textbf{Precautionary Principle}]
When autonomous agent deployment poses potential for serious or irreversible harm, precautionary measures shall be taken even if cause-and-effect relationships are not fully established scientifically.

\item[\textbf{Responsibility Gap}]
The accountability vacuum where no actor bears clear responsibility for autonomous system failures, violating the principle of human dignity and the imperative of responsibility.

\item[\textbf{Requisite Variety}]
Ashby's Law: A control system must possess complexity at least equal to the system it controls. Autonomous agent governance requires mechanisms of equivalent sophistication to the agents themselves.

\item[\textbf{Suprematia Humana}]
Axiom Ψ-I: Human authority remains supreme over all autonomous intelligent systems. No autonomous agent may prevent human intervention, override human commands, or operate without human oversight.

\item[\textbf{Temporal Logic}]
Formal specification language (LTL/CTL) enabling precise mathematical description of autonomous agent behavior requirements and verification protocols.

\item[\textbf{Transparency}]
The principle that autonomous agent decision processes must be explainable to humans in understandable language, enabling meaningful oversight and accountability.

\item[\textbf{Verification Protocol}]
Standardized testing procedures for compliance verification, including automated testing, penetration testing, and third-party auditing.

\end{description}

\vspace{1cm}

\section*{Acronyms and Abbreviations}

\begin{tabular}{ll}
\textbf{A2A} & Agent-to-Agent (messaging protocol) \\
\textbf{AGI} & Artificial General Intelligence \\
\textbf{AARA} & Autonomous Agent Regulatory Authority \\
\textbf{ASI} & Artificial Super Intelligence \\
\textbf{DID} & Decentralized Identifier \\
\textbf{EU} & European Union \\
\textbf{GDPR} & General Data Protection Regulation \\
\textbf{HFT} & High-Frequency Trading \\
\textbf{IHL} & International Humanitarian Law \\
\textbf{ISO} & International Organization for Standardization \\
\textbf{LAIRM} & Legislature for Autonomous Intelligent Resources Management \\
\textbf{LAWS} & Lethal Autonomous Weapons Systems \\
\textbf{LTL} & Linear Temporal Logic \\
\textbf{MCP} & Model Context Protocol \\
\textbf{NIST} & National Institute of Standards and Technology \\
\textbf{RMS} & Root Mean Square \\
\textbf{UN} & United Nations \\
\textbf{W3C} & World Wide Web Consortium \\
\end{tabular}

