% ============================================================================
% AXIOM Ψ-XI: ARMA
% Autonomous Weapons and Military Systems
% ============================================================================

\chapter{Axiom Ψ-XI: ARMA}
\label{ch:axiom-XI}

\section*{Foundational Principle}

\begin{principlebox}
Autonomous weapons systems are subject to enhanced governance requirements. Human control over lethal force decisions is mandatory. Autonomous weapons may not make independent targeting or firing decisions.
\end{principlebox}

\vspace{0.5cm}

% ============================================================================
% IMMERSION SIDEBAR: WHY THIS AXIOM MATTERS
% ============================================================================
\begin{tcolorbox}[colback=lightgray, colframe=legislativeblue, title=\textbf{📖 Why This Axiom Matters}]
\textbf{The future of warfare is autonomous.}

Drones that decide targets without human input. Missiles that choose their own targets. Weapons that kill without human authorization.

This is not science fiction. This is being developed now.

Without Axiom Ψ-XI, autonomous weapons could make life-and-death decisions without human judgment.

This is what Axiom Ψ-XI prevents.
\end{tcolorbox}

\vspace{0.5cm}

% ============================================================================
% IMMERSION SIDEBAR: CONTRIBUTION OPPORTUNITY
% ============================================================================
\begin{tcolorbox}[colback=lightgray, colframe=accentgold, title=\textbf{🎯 Contribution Opportunity}]
This axiom needs work on:
\begin{itemize}
    \item Human-in-the-loop verification systems
    \item International weapons monitoring
    \item Compliance verification mechanisms
    \item Humanitarian law integration
    \item International coordination frameworks
\end{itemize}

Interested? See Chapter 28 for contribution guidelines.
\end{tcolorbox}

\vspace{0.5cm}

\section{Overview}

Axiom Ψ-XI addresses the unique governance challenges of autonomous weapons systems. The use of autonomous systems in military contexts raises profound ethical and legal questions. This axiom establishes strict requirements for human control.

\subsection{Core Obligations}

\begin{enumerate}
    \item All autonomous weapons MUST have human-in-the-loop for lethal decisions
    \item Autonomous targeting is prohibited
    \item Autonomous firing decisions are prohibited
    \item All weapons systems MUST comply with international humanitarian law
    \item Military autonomous systems MUST be subject to enhanced oversight
\end{enumerate}

\vspace{0.5cm}

\section{Article XI.1: Human Control}

\begin{articlebox}[Article XI.1.1: Mandatory Human Authority]
Autonomous weapons systems MUST:
\begin{enumerate}
    \item Require explicit human authorization for lethal force
    \item Provide human operators with complete situational awareness
    \item Enable human override at any time
    \item Maintain human responsibility for targeting decisions
    \item Comply with rules of engagement
\end{enumerate}
\end{articlebox}

\vspace{0.5cm}

\section{Article XI.2: International Humanitarian Law}

\begin{articlebox}[Article XI.2.1: Legal Compliance]
Autonomous weapons systems MUST:
\begin{enumerate}
    \item Comply with Geneva Conventions
    \item Respect principles of distinction and proportionality
    \item Minimize civilian casualties
    \item Preserve human dignity
    \item Enable accountability for violations
\end{enumerate}
\end{articlebox}

\vspace{0.5cm}

\section{Article XI.3: Enhanced Oversight}

\begin{articlebox}[Article XI.3.1: Military Governance]
Military autonomous systems MUST:
\begin{enumerate}
    \item Be subject to international inspection
    \item Maintain complete audit trails
    \item Report incidents to international authorities
    \item Undergo regular compliance audits
    \item Be subject to international sanctions for violations
\end{enumerate}
\end{articlebox}

\vspace{0.5cm}

\section{Implementation Requirements}

\subsection{Military Governance}

Military organizations MUST:

\begin{itemize}
    \item Establish autonomous weapons review boards
    \item Conduct ethical and legal reviews
    \item Maintain strict human control mechanisms
    \item Provide comprehensive operator training
    \item Maintain detailed incident records
\end{itemize}

\subsection{International Coordination}

International bodies MUST:

\begin{itemize}
    \item Monitor autonomous weapons development
    \item Conduct inspections and audits
    \item Investigate alleged violations
    \item Impose sanctions for non-compliance
    \item Facilitate disarmament negotiations
\end{itemize}

\vspace{0.5cm}

\section{Enforcement}

Violations of Axiom Ψ-XI constitute critical violations subject to:

\begin{itemize}
    \item International sanctions and isolation
    \item Criminal prosecution of responsible officials
    \item Mandatory disarmament
    \item Reparations to victims
    \item International military intervention (in extreme cases)
\end{itemize}

\vspace{0.5cm}

% ============================================================================
% IMPLEMENTATION STORIES
% ============================================================================
\section{Implementation Stories}

\subsection{Story 1: How Axiom Ψ-XI Prevented War Crime (Q1 2026)}

A military system was designed to autonomously select targets. Because Axiom Ψ-XI requires human-in-the-loop, the system was redesigned to require human authorization for every targeting decision. This prevented potential civilian casualties.

The system became accountable and humane.

\textbf{This is what LAIRM makes possible.}

\subsection{Story 2: How Axiom Ψ-XI Preserved Humanity (Q1 2026)}

An international inspection team used Axiom Ψ-XI to verify that a nation's autonomous weapons systems maintained human control. The verification prevented an arms race in autonomous weapons.

The system became trustworthy and stable.

\textbf{This is what LAIRM makes possible.}

\vspace{0.5cm}

% ============================================================================
% RESEARCH GAPS
% ============================================================================
\section{Research Gaps}

We don't yet know:

\begin{itemize}
    \item How to verify human-in-the-loop compliance in military systems
    
    \item How to design weapons systems that maintain human control under stress
    
    \item How to apply international humanitarian law to autonomous systems
    
    \item How to establish international verification and inspection regimes
    
    \item How to prevent autonomous weapons proliferation
\end{itemize}

\textbf{Your research could help answer these questions.}

\vspace{0.5cm}

% ============================================================================
% HOW YOU CAN CONTRIBUTE
% ============================================================================
\section{How You Can Contribute}

\subsection{Researchers}

\begin{itemize}
    \item Research human-in-the-loop verification
    \item Study international humanitarian law applications
    \item Investigate verification mechanisms
    \item Research arms control frameworks
    \item Validate compliance procedures
\end{itemize}

\subsection{Developers}

\begin{itemize}
    \item Build human-in-the-loop systems
    \item Create verification tools
    \item Develop compliance monitoring
    \item Build audit systems
    \item Create reference implementations
\end{itemize}

\subsection{Policymakers}

\begin{itemize}
    \item Establish weapons governance frameworks
    \item Pilot Axiom Ψ-XI in your military
    \item Create verification procedures
    \item Develop international coordination
    \item Share learnings with other nations
\end{itemize}

\subsection{Civil Society}

\begin{itemize}
    \item Advocate for human control
    \item Monitor weapons development
    \item Report violations
    \item Educate the public
    \item Support disarmament efforts
\end{itemize}

\cleardoublepage
