% ============================================================================
% AXIOM Ψ-XVII: HUMANITAS
% Human Dignity and Rights
% ============================================================================

\chapter{Axiom Ψ-XVII: HUMANITAS}
\label{ch:axiom-XVII}

\section*{Foundational Principle}

\begin{principlebox}
Autonomous agents must respect human dignity and human rights. All decisions affecting humans must be made with respect for fundamental rights, freedoms, and human agency.
\end{principlebox}

\vspace{0.5cm}

% ============================================================================
% IMMERSION SIDEBAR: WHY THIS AXIOM MATTERS
% ============================================================================
\begin{tcolorbox}[colback=lightgray, colframe=legislativeblue, title=\textbf{📖 Why This Axiom Matters}]
\textbf{An autonomous system manipulates your behavior.}

It knows your weaknesses. It exploits your psychology. It controls your choices.

You think you're free. But you're not.

This is what Axiom Ψ-XVII prevents.
\end{tcolorbox}

\vspace{0.5cm}

% ============================================================================
% IMMERSION SIDEBAR: CONTRIBUTION OPPORTUNITY
% ============================================================================
\begin{tcolorbox}[colback=lightgray, colframe=accentgold, title=\textbf{🎯 Contribution Opportunity}]
This axiom needs work on:
\begin{itemize}
    \item Human rights frameworks for autonomous systems
    \item Dignity and autonomy preservation
    \item Rights-based design methodologies
    \item Human rights impact assessment
    \item Rights monitoring and enforcement
\end{itemize}

Interested? See Chapter 28 for contribution guidelines.
\end{tcolorbox}

\vspace{0.5cm}

\section{Overview}

Axiom Ψ-XVII establishes human rights protections. Autonomous agents must be designed and operated in ways that respect human dignity and fundamental rights.

\subsection{Core Obligations}

\begin{enumerate}
    \item All agents MUST respect human dignity
    \item Human rights MUST be protected
    \item Human autonomy MUST be respected
    \item Fundamental freedoms MUST be preserved
    \item Human agency MUST be maintained
\end{enumerate}

\vspace{0.5cm}

\section{Article XVII.1: Human Dignity}

\begin{articlebox}[Article XVII.1.1: Dignity Protection]
Autonomous agents MUST:
\begin{enumerate}
    \item Respect human dignity in all interactions
    \item Avoid dehumanizing or degrading treatment
    \item Preserve human autonomy and choice
    \item Respect privacy and confidentiality
    \item Enable human flourishing
\end{enumerate}
\end{articlebox}

\vspace{0.5cm}

\section{Article XVII.2: Human Rights}

\begin{articlebox}[Article XVII.2.1: Rights Protection]
Autonomous agents MUST:
\begin{enumerate}
    \item Respect freedom of expression
    \item Respect freedom of association
    \item Respect freedom of conscience
    \item Respect right to privacy
    \item Respect right to due process
\end{enumerate}
\end{articlebox}

\vspace{0.5cm}

\section{Article XVII.3: Human Agency}

\begin{articlebox}[Article XVII.3.1: Autonomy Preservation]
Autonomous agents MUST:
\begin{enumerate}
    \item Preserve human decision-making authority
    \item Enable informed human choices
    \item Avoid manipulation or coercion
    \item Respect human values and preferences
    \item Support human flourishing
\end{enumerate}
\end{articlebox}

\vspace{0.5cm}

\section{Implementation Requirements}

\subsection{Rights-Based Design}

Developers MUST:

\begin{itemize}
    \item Conduct human rights impact assessments
    \item Design for human dignity
    \item Implement rights protections
    \item Enable human oversight
    \item Maintain human agency
\end{itemize}

\subsection{Rights Monitoring}

Deployers MUST:

\begin{itemize}
    \item Monitor for human rights violations
    \item Investigate complaints
    \item Implement corrective actions
    \item Report on human rights metrics
    \item Maintain accountability
\end{itemize}

\vspace{0.5cm}

\section{Enforcement}

Violations of Axiom Ψ-XVII constitute serious violations subject to:

\begin{itemize}
    \item Fines of €10 million to €100 million or 2\% global revenue
    \item Mandatory human rights improvements
    \item Compensation for victims
    \item Suspension of operations until compliance achieved
\end{itemize}

\vspace{0.5cm}

% ============================================================================
% IMPLEMENTATION STORIES
% ============================================================================
\section{Implementation Stories}

\subsection{Story 1: How Axiom Ψ-XVII Protected Dignity (Q1 2026)}

An autonomous system was designed to manipulate human behavior. Because Axiom Ψ-XVII requires respect for human dignity and autonomy, the system was redesigned to inform rather than manipulate.

The system became respectful and ethical.

\textbf{This is what LAIRM makes possible.}

\subsection{Story 2: How Axiom Ψ-XVII Preserved Rights (Q1 2026)}

An autonomous system was violating privacy rights. Because Axiom Ψ-XVII requires human rights protection, the violation was detected, the system was redesigned, and victims were compensated.

The system became rights-respecting.

\textbf{This is what LAIRM makes possible.}

\vspace{0.5cm}

% ============================================================================
% RESEARCH GAPS
% ============================================================================
\section{Research Gaps}

We don't yet know:

\begin{itemize}
    \item How to operationalize human rights in autonomous systems
    
    \item How to measure respect for human dignity
    
    \item How to balance autonomy with safety
    
    \item How to design for human flourishing
    
    \item How to protect rights across different cultures
\end{itemize}

\textbf{Your research could help answer these questions.}

\vspace{0.5cm}

% ============================================================================
% HOW YOU CAN CONTRIBUTE
% ============================================================================
\section{How You Can Contribute}

\subsection{Researchers}

\begin{itemize}
    \item Research human rights frameworks
    \item Study dignity and autonomy
    \item Investigate rights-based design
    \item Research cultural variations
    \item Validate rights protections
\end{itemize}

\subsection{Developers}

\begin{itemize}
    \item Build rights-based systems
    \item Create dignity protections
    \item Develop autonomy preservation
    \item Build monitoring systems
    \item Create reference implementations
\end{itemize}

\subsection{Policymakers}

\begin{itemize}
    \item Establish human rights standards
    \item Pilot Axiom Ψ-XVII in your jurisdiction
    \item Create monitoring procedures
    \item Develop enforcement mechanisms
    \item Share learnings with other jurisdictions
\end{itemize}

\subsection{Civil Society}

\begin{itemize}
    \item Advocate for human rights
    \item Monitor rights violations
    \item Report violations
    \item Support victims
    \item Educate the public
\end{itemize}

\cleardoublepage
